---
title: "Integrative analysis workshop with TCGAbiolinks and ELMER"
author: "Tiago Chedraoui Silva, Simon Coetzee, Ben Berman, Houtan Noushmehr"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: true
    number_sections: no
    theme: flatly
    highlight: tango
    mathjax: null
    toc: true
    toc_float: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Integrative analysis workshop with TCGAbiolinks and ELMER}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The Cancer Genome Atlas (TCGA), The Encyclopedia of DNA Elements (ENCODE), and 
The NIH Roadmap Epigenomics Mapping Consortium (Roadmap) and other organized 
international consortia have led the explosion of sequencing based biological 
data and thereby have provided unprecedented access to the largest publicly 
available genomic, transcriptomic and epigenomic data to date. 

These projects have provided amazing opportunities for researchers to interrogate 
the epigenome of cultured cancer cell lines, normal and tumor fresh tissues with 
high genomic resolution. However, the use of such data in analyzes, comprises the 
arduous task of searching, downloading and processing them in a reproducible manner. 
Furthermore, most bioinformatics tools are designed for specific data types 
(e.g. expression, epigenetics, genomics) which provides only a partial view of 
the biological process that takes place. 

Performing an integrated analysis of molecular datasets along with clinical 
information, has been shown to improve the prognostic and predictive accuracy 
for cancer phenotypes if compared to clinical features alone. This workshop will 
focus on helping researchers to perform an integrative analysis of both molecular 
and clinical data available through the TCGA by harnessing open source packages 
within the bioconductor platform.

Participants will learn to search and download DNA methylation (epigenetic) and 
gene expression (transcription) data from the newly created 
[NCI Genomic Data Commons (GDC) portal](https://portal.gdc.cancer.gov/) 
and prepare them into a Summarized Experiment object. 
We will introduce the workflow using our recently developed TCGAbiolinks and if 
time permitted, we will highlight the Graphics User Interface version 
([TCGAbiolinksGUI](http://bioconductor.org/packages/TCGAbiolinksGUI/)). 

Another bioconductor package will also be introduced called 
[ELMER](http://bioconductor.org/packages/ELMER/) which allows 
one to identify DNA methylation changes in distal regulatory regions and correlate 
these signatures with expression of nearby genes to identify transcriptional 
targets associated with cancer. For these distal probes correlated with a gene, 
a transcription factor motif analysis is performed followed by expression 
analysis of transcription factors to infer upstream regulators.

We expect that participants of this workshop will understand the integrative 
analysis performed by using 
[TCGAbiolinks](http://bioconductor.org/packages/TCGAbiolinks/) + 
[ELMER](http://bioconductor.org/packages/ELMER/), as well as be able to execute 
it from the data acquisition process to the final interpretation of the results. 
The workshop assumes users with an intermediate level of familiarity with R, 
and basic understanding of tumor biology. 


## Installation instructions

To install this package and build the vignette run the following commands:

```{r src, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
useDevel()
biocLite("devtools")
biocLite("BioinformaticsFMRP/Bioc2017.TCGAbiolinks.ELMER",
         dependencies = TRUE, build_vignettes = TRUE)
```

## Accessing the vignette 
The vignette can be laucnhed with the following command:

```{r vignette, eval=FALSE}
library("BioinformaticsFMRP/Bioc2017.TCGAbiolinks.ELMER")
bioc2017()
```

# Session Info

```{r sessioninfo, eval=TRUE}
sessionInfo()
```
